**Policy should change to improve invited speaker diversity and retain trainees**

### 2019-09-26 submission/manuscript.Rmd
update discussion and abstract

---
title: '**Department Invited Speakers Do Not Reflect Trainee Diversity**'
csl: mbio.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: references.bib
---

@@

Running title: Invited Speaker Diversity Does Not Reflect Trainee Diversity

@@

Ada K. Hagan, Ph.D.${^1\dagger}$, Rebecca M. Pollet, Ph.D.${^1}$, and Josie Libertucci, Ph.D.${^2\dagger}$

@@

$\dagger$ To whom correspondence should be addressed: akhagan@umich.edu or libertj@mcmaster.ca

@@

2\. Department of Medicine, McMaster University, Hamilton, Ontario, Canada

@@

@@

Social role theory emphasizes the importance of "like" representation to improve feelings of self-efficacy and belonging. This is an issue for trainees in the biological sciences who are either white women or underrepresented minorities, since there is a lack of faculty who look like them in biology academic settings. One way to improve representation is through invited speaker series. In this persepective, we compare the diversity of invited speakers at one microbiology and immunology department to the diversity of their post-Baccalaurate trainees. We find that trainees are not proportionally represented by the invited speakers and provide policy suggestions and tools to address this issue.

@@

Long-standing systemic bias, sexism, and racism have contributed to the underrepresentation of many racial and ethnic groups, as well as women, in science, technology, engineering, and math (STEM) fields (NIH 2015 report, Measuring DEI 2016, [@colloquy_2012]). Organization climate and culture that supports inclusion of all individuals is essential to support the retention of historically underrepresented minorities (HURM) in the U.S., as well as non-Caucasian individuals, and women in STEM fields (Schneider, 2013). Inclusive organization culture will not suffice in retaining HURM, non-Caucasian individuals, and women without organization policy changes that support their values and goals [@Coe2019].

@@

This is a retrospective study, thus speakers were not asked for their identities at the time of visit. Instead we hand-coded proxy demographics using personal knowledge, photos, and CVs. The presenting gender of each individual was assigned using a binary system (man/woman). Due to the low number of individuals in the study, race/ethnicity demographics were split in three groups: Caucasian, Historically Under-represented Minority (HURM), and Non-Caucasian/Non-HURM (NCNH), each with a binary (yes/no) possibilty. Caucasian was assigned using the current U.S. Census definition where those of Middle Eastern, European, and Russian descent are included. HURM individuals were restricted to those with African-American, Indigenous, Alaskan/Hawaian Native, Latinx and/or Hispanic heritage. All others were placed into the NCNH group. We recognize that our proxy demographics are a limitation of the analysis and want to acknowledge that biological sex (male/female) is not always equivalent to the gender that an individual presents as (man/woman), which is also distinct from the gender(s) that an individual
self-identifies as. We also want to acknowledge that there are many other identities that are not captured in this limited analysis.

Data were analyzed and figures generated in R Statisical Software, using relevant packages [@R_software_2017; @wickham_tidyverse_2017; @cowplot; @markdown; @rmd_book; @rmd_rstudio; @knitr_2014; @knitr_2018; @lubridate; @readxl; @pdftools; @wickham_scales_2018; @neuwirth_rcolorbrewer_2014].

@@

trainee_summary <- tidy_trainee %>% 
  select(-Gender) %>% 
  filter(value == "y") %>% distinct_() %>% 
  group_by(demographic) %>% summarise(n = n()) %>% 
  mutate(prop = get_percent(n, sum(.$n)))
```
To understand the representation of women, we compared the proportion of women in each academic role. At the trainee level, more than half of students and postdoctoral fellows were women. That dropped to `r gender_combined[[1,4]]`% of faculty hosts and `r gender_combined[[3,4]]`% of the invited speakers (Fig. 1A). Of `r sum(lect_summary$n)` lectureships over the five year period, `r lect_summary[4,4]`% were awarded to women. The proportion of women as faculty hosts and speakers is equivalent to global estimates that 40% of microbiologists are women, with a slightly lower representation of women in lectureships [@allagnat_gender_2017].

Our analysis identified an over-representation of Caucasian individuals as hosting faculty and invited speakers (80% each), relative to the proportion of Caucasian trainees, which was 55% (Fig. 1B). We also observed declines in the representation of HURM and NCNH faculty and speakers relative to the trainees (Fig 1B). HURM trainees made up 11% of the department, on track with the 11% of microbiology and immunology doctorates awarded in 2017 [@nsf_survey_2017]. However, only 8.5% of invited speakers, and none of the hosting faculty, were HURM scientists. NCNH trainees were 34% of department students and postdocs (22% of microbiology and immunology doctorates in 2017), but only 19% of hosting faculty and 10.5% of invited speakers [@nsf_survey_2017].

The more prestigous invited speaker lectureships were also dominated by Caucasian scientists, who comprised `r get_percent(sum(cauc_lect$sum), sum(lect_summary$n))`% of those awarded (Fig. 1C). HURM and NCNH scientists were awarded `r lect_summary[[2,3]]` and `r lect_summary[[3,3]]` lectureships, respectively. Because the intersection of identities can compound biases and outcomes, we further examined the lectureships by gender and race/ethnicity status. Caucasian men and women accounted for `r lect_summary[[1,4]]`% and `r lect_summary[[4,4]]`% of the lectureships, respectively. Just `r 100 - get_percent(sum(cauc_lect$sum), sum(lect_summary$n))`% of lectureships were held by non-Caucasian men while none were held by non-Caucasian women (Fig. 1D).

@@

[Add paragraph summarizing findings ] According to social role theory, this department is not adequately serving their non-Caucasian trainees.

Several papers have investigated the representation of women at scientific conferences, however, we have only identified one that focused on invited speakers at universities [@nittrouer_gender_2018]. In their study, Nittrouer et al, examined 3,652 talks at 50 U.S. institutions in 2013 - 2014 and found that women faculty are less likely to be invited speakers, despite similar acceptance rates. These results suggest that women faculty are less often invited as speakers. We have not been able to identify any other publications examining scientific speaker diversity beyond gender. This seems to be the first. This is concerning since conclusions drawn from gender-based studies are often framed, and considered, to be applicable to other marginalized groups (e.g., HURM). This is a flawed assumption. While there is no doubt some overlap, each group remains marginalized due to a unique complex set of factors that cannot always be solved by gender-based solutions. US-serving institutions, such as the University of Michigan have a particular responsibility to the historically suppressed populations included in our definition of HURMs. We therefore implore US institutions to apply this framing to their discussions and research.

@@

While HURM and NCNH share some experiences, differences including varying rates of hiring and tenure promotion mean unique considerations are important for inclusion of each group [@Fang2000]. For instance, a major barrier to inclusion of HURM faculty at similar proportions to HURM trainees is the low transition rate of scientists from HURM backgrounds to faculty positions and the associated low proportion of HURM faculty [@Gibbs2016]. The proportion of HURM faculty at the Assistant and Associate Professor level is currently higher than at Full Professor so it will be difficult to increase speaker diversity if early-career researchers are not being considered [@whittaker_retention_2015]. Increased performance expectations and patterns of exclusions are consistent themes in studies characterizing the HURM faculty experience [@pololi_race_2010; @hassouneh_experiences_2014]. Therefore, inclusion of HURM faculty in seminar series is likely essential to increasing the number of HURM Associate and Full Professors. Even when HURM speaker rates match the proportion of HURM faculty employment, HURM trainees will be represented at a significantly higher proportion. Inclusion of HURM faculty in these seminarseries is just one aspect of larger institutional change that is needed [@johnson_msphere_2019].

## Instituting policy change within the Department of Microbiology and Immunology

In an attempt to promote inclusion within the Dept. M&I these data were presented to faculty members and the dept chair (Harry Mobley). Since, trainee demographics were not represented within seminar speaker demographics over the past 5 years, we proposed a policy change as to how seminar speakers were being invited. This policy change included switching from PI invited to lab invited in an attempt to allow trainees to choose a speaker that best represented themselves.

Implict biases that affect perceptions of marginalized groups are the primary issue, but we must acknowledge that it is not always possible to identify members of historically under-served communities. For instance, after data analysis, we learned that at least one speaker in our data set should have been categorized as a HURM instead of Caucasian, but it wasn't readily apparent from their internet presence or CV. This limitation makes two important points: that percieved identity often plays a larger role than self-identification, and that we need better tools to identify members of marginalized groups. So in addition to our policy change, we also created resources that allow scientists to self-identify and provide host faculty with more diverse choices.

@@

To help address this issue, we make some suggestions (Table 1) and have developed a resource to identify scientists who are members of marginalized and/or historically under-served groups. Motivated by a lack of such resources and inspired by similar resources--DiversifyEEB and DiversifyChemistry--we created DiversifyMicrobiology and DiversifyImmunology. These resources are a tool for symposium organziers, award committees, search committees and other scientists to identify individuals to diversify their pools. Additionally, we have built these as a tool for use by other fields and organizations to create their own lists. Since these lists are compiled by self-nomination, we can ensure that only scientists comfortable revealing their marginalized identities are included.

@@

The website provides an interface to the Google forms and spreadsheets with template pages for viewing the list, adding a name to the list, and finding additional resources. Importantly, our website creation tool is hosted for free by GitHub, which provides a free website for each GitHub organization. Basic tools and skills required to set up a Diversify site include knowledge of, or experience with, the version control tool git, the webtool GitHub, and a text editor. A tutorial in the DiversifyMicrobiology repository on GitHub provides links to these resources and instructions for adapting the tool to your own field.

@@

To increase the retention of white women and HURMS in STEM, they must also be represented as experts. However, the invited speaker diversity at one department does not represent the diversity of trainees. There is a lack of research on invited speakers examining factors other than gender. To facilite the identification and recruitment of individuals in these historically under-served groups, we have built a tool to create self-nominated, field-specific lists.

@@

We thank Drs. Beth Moore and Harry Mobley and the Department of Microbiology & Immunology, University of Michigan for their input and financial support that enabled publication of our manuscript. We would also like to acknowledge Nick Lesniak and Dr. Ariangela Kozick for their comments and suggestions.

## Author Contributions
A.K.H. collected the data, assigned demographics, analyzed the data, and created the website. R.M.P. created the Google lists, forms, and website content and the description of their maintenance. J.L. wrote the introduction and provided conceptual advice. A.K.H. and J.L facilited the policy change to the UM Department of Microbiology and Immunology. All authors contributed to preparing the final manuscript.

@@

Use a list & \makecell[l]{Many lists of scientists from \\under-represented and underserved \\groups are available} &  \makecell[l]{https://DiversifyMicrobiology.\\github.io/resources}\\ \hline

@@

![**The demographics of invited speakers, hosting faculty, and trainees.** A) The proportion of women in each academic role. B) The proportion of each academic role represented by individuals that are Caucasian (left), Historically Underrepresented Minorities (HURM, center) or International (right). C-D)The percent of lectureships awarded to individuals that are C) Caucasian, HURM, International or D) Caucasian or Non-Caucasian by gender.](Figure_1.png)

>>>>

---
title: '**Policy should change to improve invited speaker diversity and retain trainees**'
csl: mbio.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: references.bib
---

@@

Running title: Suggestions to improve invited speaker diversity

@@

Ada K. Hagan, Ph.D.${^1\dagger}$, Rebecca M. Pollet, Ph.D.${^1}$, and Josie Libertucci, Ph.D.${^2}$

@@

$\dagger$ To whom correspondence should be addressed: akhagan@umich.edu

@@

2\. Farncombe Family Digestive Health Research Institute, Department of Medicine, McMaster University, Hamilton, Ontario, Canada

@@

Financial support: Department of Microbiology & Immunology, University of Michigan

@@

The biological sciences have a problem retaining white women and underrepresented minorities in academia. Despite increases in the representation of these groups in faculty candidate pools, they are still underrepresented at the faculty level, particularly at the Full professor level. Faculty are often invited to departments at other universities to present on their research, a highly valued offer that provides an opportunity for collaborations and networking. This is also an opportunity for trainees to interact with additional faculty from underrepresented groups. In this manuscript, we compare invited-speaker demographics to the current trainee demographics and find that trainees are not proportionally represented by speakers invited to the department. Our investigation prompted changes in policy for how invited speakers are selected in the future to invite a more diverse group of scientists. To facilitate this process, we developed a set of tips and a web-based resource that allows scientists, committees, and moderators to identify members of under-served groups. These resources can be easily adapted by other fields or sub-fields to promote inclusion and diversity at seminar series’, conferences, and colloquia.

@@

**waiting for updated version**

Long-standing systemic bias, sexism, and racism have contributed to the under-representation of many racial and ethnic groups, as well as women, in science, technology, engineering, and math (STEM) fields (NIH 2015 report, Measuring DEI 2016, [@colloquy_2012]). Organization climate and culture that supports inclusion of all individuals is essential to support the retention of historically underrepresented minorities (HURM) in the U.S., as well as non-Caucasian individuals, and women in STEM fields (Schneider, 2013). Inclusive organization culture will not suffice in retaining HURM, non-Caucasian individuals, and women without organization policy changes that support their values and goals [@Coe2019].

@@

This is a retrospective study, thus speakers were not asked for their identities at the time of visit. Instead we hand-coded proxy demographics using personal knowledge, photos, and CVs. The presenting gender of each individual was assigned using a binary system (man/woman). Due to the low number of individuals in the study, race/ethnicity demographics were split in three groups: Caucasian, Historically Under-represented Minority (HURM), and Non-Caucasian/Non-HURM (NCNH), each with a binary (yes/no) possibility. Caucasian was assigned using the current U.S. Census definition where those of Middle Eastern, European, and Russian descent are included. HURM individuals were restricted to those with African-American, Indigenous, Alaskan/Hawaiian Native, Latinx and/or Hispanic heritage. All others were placed into the NCNH group. We recognize that our proxy demographics are a limitation of the analysis and want to acknowledge that biological sex (male/female) is not always equivalent to the gender that an individual presents as (man/woman), which is also distinct from the gender(s) that an individual
self-identifies as. We also want to acknowledge that there are many other identities that are not captured in this limited analysis.

Data were analyzed and figures generated in R Statistical Software, using relevant packages [@R_software_2017; @wickham_tidyverse_2017; @cowplot; @markdown; @rmd_book; @rmd_rstudio; @knitr_2014; @knitr_2018; @lubridate; @readxl; @pdftools; @wickham_scales_2018; @neuwirth_rcolorbrewer_2014].

@@

trainee_summary <- tidy_trainee %>% 
  select(-Gender) %>% 
  filter(value == "y") %>% distinct_() %>% 
  group_by(demographic) %>% summarise(n = n()) %>% 
  mutate(prop = get_percent(n, sum(.$n)))
```
To understand the representation of women, we compared the proportion of women in each academic role. At the trainee level, more than half of students and postdoctoral fellows were women. That dropped to `r gender_combined[[1,4]]`% of faculty hosts and `r gender_combined[[3,4]]`% of the invited speakers (Fig. 1A). Of `r sum(lect_summary$n)` lectureships over the five year period, `r lect_summary[4,4]`% were awarded to women.

Our analysis identified an over-representation of Caucasian individuals as hosting faculty and invited speakers (80% each), relative to the proportion of Caucasian trainees, which was 55% (Fig. 1B). We also observed declines in the representation of HURM and NCNH faculty and speakers relative to the trainees (Fig 1B). HURM trainees made up 11% of the department, on track with the 11% of microbiology and immunology doctorates awarded in 2017 [@nsf_survey_2017]. However, only 8.5% of invited speakers, and none of the hosting faculty, were HURM scientists. NCNH trainees were 34% of department students and postdoc (22% of microbiology and immunology doctorates in 2017), but only 19% of hosting faculty and 10.5% of invited speakers [@nsf_survey_2017].

The more prestigious invited speaker lectureships were also dominated by Caucasian scientists, who comprised `r get_percent(sum(cauc_lect$sum), sum(lect_summary$n))`% of those awarded (Fig. 1C). HURM and NCNH scientists were awarded `r lect_summary[[2,3]]` and `r lect_summary[[3,3]]` lectureships, respectively. Because the intersection of identities can compound biases and outcomes, we further examined the lectureships by gender and race/ethnicity status. Caucasian men and women accounted for `r lect_summary[[1,4]]`% and `r lect_summary[[4,4]]`% of the lectureships, respectively. Just `r 100 - get_percent(sum(cauc_lect$sum), sum(lect_summary$n))`% of lectureships were held by non-Caucasian men while none were held by non-Caucasian women (Fig. 1D).

@@

In our study population, the proportion of women as faculty hosts and speakers is equivalent to global estimates that 40% of microbiologists are women, with a slightly lower representation of women in lectureships [@allagnat_gender_2017]. In that regard, women are over-represented as graduate students and postdoctoral fellows in the department. Caucasian scientists are over-represented as host faculty and invited speakers, compared to their presence as trainees, particularly when lectureships were considered. Conversely, the proportion of HURM and NCNH invited speakers were under-representative of the trainee populations for each group. Thus, according to social role theory, this department is not adequately serving their non-Caucasian trainees.

Several papers have investigated the representation of women at scientific conferences, however, we have only identified one that focused on invited speakers at universities [@nittrouer_gender_2018]. In their study, Nittrouer et al, examined 3,652 talks at 50 U.S. institutions in 2013 - 2014 and found that women faculty are less likely to be invited speakers, despite similar acceptance rates. These results suggest that women faculty are less often invited as speakers. We have not been able to identify any other publications examining scientific speaker diversity beyond gender. This seems to be the first, which is concerning since conclusions drawn from gender-based studies are often framed, and considered, to be applicable to other marginalized groups (e.g., HURM). This is a flawed assumption. While there is no doubt some overlap, each group remains marginalized due to a unique complex set of factors that cannot always be solved by gender-based solutions. US-serving institutions, such as the University of Michigan have a particular responsibility to the historically suppressed populations included in our definition of HURMs. We therefore implore US institutions to apply this framing to their discussions and research.

@@

While HURM and NCNH share some experiences, differences including varying rates of hiring and tenure promotion mean unique considerations are important for inclusion of each group [@Fang2000]. For instance, a major barrier to inclusion of HURM faculty at similar proportions to HURM trainees is the low transition rate of scientists from HURM backgrounds to faculty positions and the associated low proportion of HURM faculty [@Gibbs2016]. The proportion of HURM faculty at the Assistant and Associate Professor level is currently higher than at Full Professor so it will be difficult to increase speaker diversity if early-career researchers are not being considered [@whittaker_retention_2015]. Increased performance expectations and patterns of exclusions are consistent themes in studies characterizing the HURM faculty experience [@pololi_race_2010; @hassouneh_experiences_2014]. Therefore, inclusion of HURM faculty in seminar series is likely essential to increasing the number of HURM Associate and Full Professors. Even when HURM speaker rates match the proportion of HURM faculty employment, HURM trainees will be represented at a significantly higher proportion. Inclusion of HURM faculty in these seminar series is just one aspect of larger institutional change that is needed [@johnson_msphere_2019].

## Instituting Policy Change

In an attempt to promote inclusion within the Department of Microbiology and Immunology at the University of Michigan, these data were presented to faculty members and the department chair (Dr. Mobley). Since trainee demographics were not represented within seminar speaker demographics over the past five years, we proposed a policy change as to how seminar speakers were being invited. One suggestion was to switch from faculty-invited to lab-invited speakers in an attempt to allow trainees to choose a speaker that best represented themselves (Table 1).

The implicit biases that affect perceptions of marginalized groups are an issue, but we must acknowledge that it is not always possible to identify members of historically under-served communities. For instance, after data analysis, we learned that at least one speaker in our data set should have been categorized as a HURM instead of Caucasian, but it wasn't readily apparent from their internet presence or CV. This limitation makes two important points: that perceived identity often plays a larger role than self-identification, and that we need better tools to identify members of marginalized groups. Another policy suggestion is for departments to invite their speakers to spend a time discussing their personal journeys through science, in addition to their scientific stories (Table 1). This would enable those who wish, feel welcomed to discuss how their identity(ies) interacted with their careers. In addition to these suggestions for policy change, we have created resources that allow scientists to self-identify as under-served groups and thus provide host faculty with more diverse choices (Table 1).

@@

Motivated by a lack of resources to identify scientists who are members of marginalized and/or historically under-served groups, and inspired by resources in other fields--DiversifyEEB and DiversifyChemistry--we created DiversifyMicrobiology and DiversifyImmunology. These resources are a tool for symposium organizers, award committees, search committees, and other scientists to identify individuals to diversify their pools. Additionally, we have built these as a template to be used as a tool for other fields and organizations that wish to create their own lists. Since these lists are compiled by self-nomination, we can ensure that only scientists comfortable revealing their marginalized identities are included.

@@

The website provides an interface to the Google forms and spreadsheets with template pages for viewing the list, adding a name to the list, and finding additional resources. Importantly, our website creation tool is hosted for free by GitHub, which provides a free website for each GitHub organization. Basic tools and skills required to set up a Diversify site include knowledge of, or experience with, the version control tool git, the web-tool GitHub, and a text editor. A tutorial in the DiversifyMicrobiology repository on GitHub provides links to these resources and instructions for adapting the tool to your own field.

@@

To increase the retention of white women and HURM trainees in the biological sciences, they must also be represented as experts. However, the invited speaker diversity at one department does not represent the diversity of trainees. To facilitate the identification and recruitment of individuals in these historically under-served groups, we have built a tool to create self-nominated, field-specific lists.

@@

We thank Drs. Beth Moore and Harry Mobley and the Department of Microbiology & Immunology, University of Michigan for their input and financial support that enabled publication of our manuscript. We thank Bonnie Krey and former speaker series coordinators Drs. Nicole Koropatkin and Kathy Spindler for access to invited speaker data. We would also like to acknowledge and thank Nick Lesniak and Dr. Ariangela Kozick for their comments and suggestions.

## Author Contributions
A.K.H. collected the data, assigned demographics, analyzed the data, created the website, and wrote the methods and results. R.M.P. created the Google lists, forms, and website content and the description of their maintenance. J.L. wrote the introduction and provided conceptual advice. A.K.H. and J.L facilitated the policy change to the UM Department of Microbiology and Immunology. All authors contributed to preparing the final manuscript.

@@

Use a list & \makecell[l]{Many lists of scientists from \\under-represented and under-served \\groups are available} &  \makecell[l]{https://DiversifyMicrobiology.\\github.io/resources}\\ \hline

@@

![**The demographics of invited speakers, hosting faculty, and trainees.** A) The proportion of women in each academic role. B) The proportion of each academic role represented by individuals that are Caucasian (left), Historically Underrepresented Minorities (HURM, center) or Non-Caucasian/Non-HURM (NCNH, right). C-D)The percent of lectureships awarded to individuals that are C) Caucasian, HURM, or NCNH and D) Caucasian, HURM, or NCNH by gender.](Figure_1.png)
